Rare-event sampling applied to the simulation of extreme mechanical efforts exerted by a turbulent flow on a bluff body

### 2019-05-22 draft_p2.tex
Add references to issues to Emmanuel's comments

\begin{abstract}
	\EL{ ABSTRACT TO BE REVIEWED}
This study evaluates the relevance of  rare-event sampling to optimize the numerical simulation of extreme mechanical efforts exerted by a turbulent flow impinging onto a bluff body. The main idea is here to replace a long simulation by a set of much shorter simulations (running in parallel) with dynamics that are replicated or abandoned in order to sample large-amplitude events more frequently. 
%
Such technique has already proved its efficiency for simulating extreme events in {...} with significant run-time savings.
Application to fluid-structure interaction problems remains mainly open. 
%
The drag experienced by a squared obstacle placed in a turbulent flow (in two dimensions) is taken as a representative case study to investigate the performance of two major sampling algorithms, namely {...} 
Practical evidence is given that these algorithms allow for the correct sampling of rare large-amplitude drag events and the estimation of return times with a reduced computational effort. However, {limitations and pitfalls}. 
% Importantly, such techniques also provide direct access to the flow scenario yielding extreme events. In the present case, it is found that {<blocking effect>}.
\end{abstract}

@@

%
% next two paragraphs may be too technical -- Find a way to summarize them in a few sentences accessible to a broad audience in fluid mechanics? Notion of action will be vague for many readers
%
% this paragraph would certainly need to be reshaped!
%
In fluid turbulence, rare-event sampling has been approached mainly from simplified dynamics such as the one-dimensional Burgers' equation with a stochastic forcing \cite{bec_burgers_2007}. In that case, dynamics may be accounted for through an action that can be identified as analogous to a Gibbs measure for a system at equilibrium. This latter can be sampled by using a Markov chain Monte-Carlo  algorithm \cite{duben_monte_2008,mesterhazy_anomalous_2011,mesterhazy_lattice_2013} that provides a framework for rare-event sampling.
%
An alternative approach is based on \emph{instanton} \cite{grafke_instanton_2015}. For stochastically driven systems in the limit of weak noise, it can be proved that the probability of extreme events is dominated by the minima of the action \cite{freidlin2013random} -- the trajectories that achieve such minima being referred to as instantons \cite{gurarie_instantons_1996}. Therefore, suitable numerical schemes can  be used to minimize the action and evaluate instantons as well as the related probabilities of rare events \cite{chernykh_large_2001,grafke_instanton_2013,grafke_relevance_2015,grigorio_instantons_2017}.
%
An obvious drawback of the aforementioned approaches is their limitation to stochastically driven dynamics which, in addition, must be simple enough so that an action may be derived and sampled, or minimized. 
%
%
%Whilst Burgers' turbulence provides an interesting model framework for the investigation of the properties of small scale turbulence, it is of little relevance to more realistic situations, such as industrial or environmental flows.
%
In this paper, a more general approach is investigated. It relies on sampling algorithms connected to the simulation of the actual dynamics itself, \textit{i.e.} applying directly to the trajectories in the dynamic space.
% dedicated to dynamical systems for which an action cannot be easily derived or minimized,
%    
%
Even though such ideas trace back to the early 1950's, they have received ever-growing interest over the last 20 years with successful applications in various domains such as chemistry \cite{van_erp_elaborating_2005,escobedo_transition_2009,teo_adaptive_2016}, biophysics \cite{huber_weighted-ensemble_1996,zuckerman2017weighted,bolhuis2005kinetic}, nuclear physics \cite{louvin2017}, nonlinear dynamical systems \cite{tailleur_probing_2007} or communication networks simulation \cite{villen-altamirano_restart:_1994}.
%
% already mentioned in the beginning? This sentence is misleading.
%
%However, a large body of algorithms are limited to systems at equilibrium, for which the stationary measure can be sampled by means of Monte Carlo methods. 
%
An original contribution of the present study is certainly the application of rare-event sampling algorithms in the context of far-from-equilibrium dynamics with an irreducible very large number of degrees of freedom. 
Let us remark that most of previous successful applications involve dynamical systems at statistical equilibrium and low-dimensional attractors. \EL{is it TRUE?}
% 
Two different algorithms suitable for out-of-equilibrium dynamics are considered and compared. Namely, the Adaptive Multilevel Splitting algorithm and the Giardina-Kurchan-Tailleur-Lecomte algorithm.
%

@@

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{ecoulement_typique/ecoulement_typique.png}
	\caption{Snapshots of the vorticity related to typical drag fluctuations (within one standard deviation) over a time interval of length $\tau_c \simeq 4\tau_0$; $\tau_c$ will be identified later as the correlation time of the drag signal.
	The vorticity is given in lattice units. \EL{THE COLORMAP IS REVERSED IN THE WHOLE PAPER: BLUE SHOULD BE NEGATIVE AND RED POSITIVE}}
	\label{fig:typical_vorticity}
\end{figure}

@@

\begin{figure}
	\centering
	%\includegraphics[width=\linewidth]{timeseries_extrms_AVG/timeseries_extrms_AVG}
	\caption{Instantaneous drag signals $f_d(t)$ corresponding to the highest fluctuations of the time-averaged drag $F_T$ with averaging period $T=10~\tau_c$ and $\tau_c$ being the correlation time of the drag. $\sigma_T$ denotes the standard deviation of $F_T$. \EL{FIGURE IS MISSING}}
	\label{fig:extreme_avg}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\linewidth]{illustr_AMS/figure_AMS}
  \caption{\label{fig:illustr_AMS} Selection-mutation procedure of the \ac{ams} algorithm to compute the probability that a real-valued observable $A(\mathbf{x}(t))$ reaches values larger than $Q$ throughout a trajectory of duration $T_a$.\EL{QUID of the score function mentioned in the text: is the observable?} The initial ensemble of blue trajectories (associated with the time evolution of the observable $A$) is first generated. These trajectories are independent and can be simulated in parallel. On the basis of the respective maxima $Q_1$, $Q_2$ and $Q_3$, the trajectory with the lowest maximum is discarded in the ensemble (dashed blue line). Among the two remaining trajectories, trajectory 3 is chosen randomly and copied until it reaches the value $Q_1$. It is then freely simulated from the branching point to the final time $T_a$. In case of deterministic dynamics, a small perturbation is introduced at the branching to separate the trajectories. This procedure is iterated $J$ times or until all trajectories exceed a fixed threshold $Q$.}
\end{figure}

@@

% Application to instantaneous drag
%
% Can the \ac{tams} algorithm achieve similar performance when applied to complex chaotic dynamics, such as turbulent flows around objects ?
% To answer this question, we consider again the two-dimensional test flow introduced in section~\ref{sec:test_flow}.
%
Our aim is now to sample flow evolution that exhibit extreme fluctuations of the drag $f_d$ acting on the square obstacle.
In contrast with the \acl{ou} process, the phase space is here higlhy-dimensional with intrinsically chaotic dynamics.
In that situation, the choice of the score function $\xi (\mathbf{x}(t),t)$ is no longer straightforward.
However, a natural simple choice for which we will opt in this study is certainly $\xi (\mathbf{x}(t),t) = f_d(t)$. 
\TL{Let us stress here that the flow dynamics are deterministic.
As a result, resampled trajectories do not differ from the parent trajectory their are resampled
from.
A workaround is to introduce a slight perturbation in the state of the flow at the branching point (see figure~\ref{fig:illustr_AMS}).
Because of the flow chaoticity, resampled trajectories will separate from their parent over the \textit{Lyapunov timescale} $\tau_L$.}\EL{Freddy: provide reference for Lyapunov timescale.}

@@

The difference between the typical duration of drag fluctuations $\tau_c$ and the Lyapunov timescale $\tau_L$ may be heuristically associated with the ``convective nature'' of the flow.
%
As discussed previously, the duration of extreme fluctuations of the drag are closely related to the sweeping time of the flow past the obstacle, and therefore to the mean velocity $U$. On the contrary, the Lyapunov timescale is rather associated with the intrinsic evolution of turbulent fluctuations in the reference frame of the mean flow, \textit{i.e.} with the fluctuating velocity $u_{rms}$. The ratio $u_{rms}/U < 1$ in our case of grid-generated turbulence implies that $\tau_L > \tau_c$. \EL{does it make sense to you?}

@@

The sampling of extreme fluctuations of the time-averaged drag $F_T$ is now examined.
The \ac{tams} algorithm could be used in the same way as before by choosing $F_T$ as score function. However, this would lead to the same pitfall.
% the re-sampling fails to efficiently generate new higher fluctuations of $F_T$.
%
For a time-averaged observable,  an alternative approach is provided by the \acf{gktl} algorithm \cite{giardina_direct_2006,tailleur_probing_2007,giardina_simulating_2011}.
%
Similarly to the \ac{tams} algorithm, the \ac{gktl} algorithm relies on the simulation of an ensemble of trajectories.
In this case, however, the trajectories interact dynamically. 
%
At regular time intervals, some elements of the ensemble are killed and others are cloned according to a weight that depends on the history of the replica.
%
The weights are chosen so that, after several iterations of the algorithm, the trajectories generated are distributed according to a biased probability distribution in favor of trajectories with large values of the (time-averaged) observable of interest.
%
The \ac{gktl} algorithm belongs to a family of algorithms called ``go with the winners'' \cite{aldous1994go,grassberger2002go}.
Similar ideas have been applied in a wide range of fields over the past 50 years, under different names \EL{donner des exemples}, depending of the specific application domain~\cite{moral_feynman-kac_2004}.
The application of a go-with-the-winners approach to the computation of large deviations in non-equilibrium systems has first been proposed in 2006 \cite{giardina_direct_2006}.
Over the past ten years, it has successfully been applied to investigate rare events in both stochastic \cite{giardina_direct_2006,lecomte_numerical_2007,garrahan2007dynamical} and deterministic systems \cite{giardina_direct_2006,tailleur_probing_2007}.
% It was originally designed to compute large deviations rate functions of time-averaged dynamical observables, that describe the probability density of both typical of rare events in the limit of large
% averaging time~\cite{ref_large_dev}.
% By contrast with the \ac{ams} and \ac{tams}, the \ac{gktl} algorithm does not follow the strategy of splitting algorithms.
% Instead, it implements \textit{importance sampling}, \textit{i.e.} sampling a modified distribution that is biased towards the rare events of interest.
% The idea of importance sampling is very general and was used in many different contexts (see e.g.~\cite{Berg1992,Hartmann2002} and the general references~\cite{Bucklew2004,RubinoTuffin2009}). 
% The \ac{gktl} algorithm performs importance sampling in the space of trajectories, which is relevant for out-of-equilibrium systems.

@@

Let 
$\mathbb{P}_{0}\left(\left\{ \mathbf{X}(t)\right\} _{0\leq t\leq T_{a}} = \left\{ \mathbf{x}(t)\right\} _{0\leq t\leq T_{a}}\right)$ 
refer formally to the probability of the trajectory 
$\left\{ \mathbf{x}(t)\right\} _{0\leq t\leq T_{a}}$ of our dynamical system.
%
% $\mathbb{P}_{k}\left(\left\{ \mathbf{X}(t)\right\} _{0\leq t\leq T_{a}} = \left\{ \mathbf{x}(t)\right\} _{0\leq t\leq T_{a}} \right)$ 
%
By construction, the probability to obtain the same trajectory by resorting to the previous cloning procedure is given by
%
\begin{align}
\mathbb{P}_{k}\left(\left\{ \mathbf{X}(t)\right\} _{0\leq t\leq T_{a}}=\left\{ \mathbf{x}(t)\right\} _{0\leq t\leq T_{a}}\right) &\underset{N\rightarrow\infty}{\sim} \frac{e^{k\int_{0}^{T_{a}}A(\mathbf{x}(t))dt}}{Z(k,T_a)}\mathbb{\mathbb{P}}_{0}\left(\left\{ \mathbf{X}(t)\right\} _{0\leq t\leq T_{a}}=\left\{ \mathbf{x}(t)\right\} _{0\leq t\leq T_{a}}\right)
\label{eq:Biased_Path_Approximation}
\end{align}
where the normalization factor is given by $Z(k,T_a)=\mathbb{E}_{0}\left[e^{k\int_{0}^{T_{a}}A(\mathbf{X}(t))dt}\right]$ with $\mathbb{E}_{0}$ being the expectation value with respect to $\mathbb{P}_{0}$.
%
% obvious? 
%and $\underset{N\rightarrow\infty}{\sim}$ meaning that this is true only asymptotically for large $N$.
%
%
%The typical error is of order $1/\sqrt{N}$ when evaluating averages over observables.
%
Eq.~\eqref{eq:Biased_Path_Approximation} relies on the mean-field approximation
\begin{equation}
R_{1}=\frac{1}{N}\sum_{n=1}^{N}e^{k\int_{t_0}^{t_{1}}A(\mathbf{X}_{n}(t))dt}\underset{N\rightarrow\infty}{\sim} Z(k,t_1)= \mathbb{E}_{0}\left[e^{k\int_{t_0}^{t_{1}}A(\mathbf{X}(t))dt}\right]
\label{eq:Mean_Field_Approximation}
\end{equation}
The typical relative error related to this approximation can be shown of being of order $1/\sqrt{N}$\cite{DelMoralBook,DelMoral2013} for a family of rare-event algorithms including the \ac{gktl} algorithm.
which by induction (and assuming similar relations for the successive $R_i$'s) eventually leads  to~\cite{giardina_direct_2006,giardina_simulating_2011}
\begin{equation}
\prod_{i=1}^{T_{a}/\tau}R_{i}\underset{N\rightarrow\infty}{\sim} Z(k,T_a) =\mathbb{E}_{0}\left[e^{k\int_{0}^{T_a}A(\mathbf{X}(t))dt}\right].\label{eq:Estimate_Lambda}
\end{equation}
\EL{IS IT TRUE?}
The validity of this mean-field approximation and the 
%
%
The killed trajectories, \emph{i.e.} with a weight smaller than $1$ \EL{true?}, are discarded from the statistics.
%
% no because the trajectories are cloned
%
% Starting from the final $N$ trajectories at time $T_{a}$, one goes backwards in time through the selection events attaching to each piece of trajectory its ancestor.
%
Eventually, an effective ensemble of $N$ trajectories for $0 \leq t \leq T_{a}$ is obtained, distributed according to $\mathbb{P}_{k}$.
All trajectories are solutions of the dynamical system since we have not modified the dynamics, but only sampled trajectories according to the distribution $\mathbb{P}_{k}$ rather than according to the distribution $\mathbb{P}_{0}$.
\EL{quid the small perturbation to separated the clones? comment required}
%
The sampled trajectories can be used to compute the statistical properties of any observable with respect to the distribution $\mathbb{P}_{0}$ from the distribution $\mathbb{P}_{k}$ by using Eq.~\eqref{eq:Biased_Path_Approximation}.
%
More details concerning the implementation of the algorithm and the corresponding data analysis are provided in \cite{lestang:tel-01974316,brewer2018efficient}.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{IS_GKTL/IS_GKTL}
	\caption{\label{fig:IS_GKTL} Rare-event sampling of the (zero-mean) time-averaged drag $\tilde F_T = F_T - \overline{F_T}$ with $T=10~\tau_c$; $\tau_c$ is the correlation time of the instantaneous drag. The shaded \ac{pdf}s are estimated from the biased ensemble resulting from the \ac{gktl} algorithm applied to $16384$ trajectories of duration $T_a=T$ with the cloning period $\tau=\tau_c/2$.
	The dashed line denotes the unbiased \ac{pdf} of $\tilde F_T$, \textit{i.e.} from direct sampling or with $k=0$ (no bias) in the \ac{gktl} algorithm.
\EL{$F_T-\overline{F_T}$ in the figure!}}
\end{figure}

@@

\EL{rajouter le calcul des temps de retour par gktl pour etoffer un peu la partie du papier sur gktl -- il faudrait montrer un resultat qui montre que l'on obtient la bonne statistique}

@@

\EL{CONCLUSION TO BE REVIEWED}

>>>>

\begin{abstract}
	\EL{ ABSTRACT TO BE REVIEWED \#8}
This study evaluates the relevance of  rare-event sampling to optimize the numerical simulation of extreme mechanical efforts exerted by a turbulent flow impinging onto a bluff body. The main idea is here to replace a long simulation by a set of much shorter simulations (running in parallel) with dynamics that are replicated or abandoned in order to sample large-amplitude events more frequently. 
%
Such technique has already proved its efficiency for simulating extreme events in {...} with significant run-time savings.
Application to fluid-structure interaction problems remains mainly open. 
%
The drag experienced by a squared obstacle placed in a turbulent flow (in two dimensions) is taken as a representative case study to investigate the performance of two major sampling algorithms, namely {...} 
Practical evidence is given that these algorithms allow for the correct sampling of rare large-amplitude drag events and the estimation of return times with a reduced computational effort. However, {limitations and pitfalls}. 
% Importantly, such techniques also provide direct access to the flow scenario yielding extreme events. In the present case, it is found that {<blocking effect>}.
\end{abstract}

@@

%
% next two paragraphs may be too technical -- Find a way to summarize them in a few sentences accessible to a broad audience in fluid mechanics? Notion of action will be vague for many readers
%
% this paragraph would certainly need to be reshaped!
%
In fluid turbulence, rare-event sampling has been approached mainly from simplified dynamics such as the one-dimensional Burgers' equation with a stochastic forcing \cite{bec_burgers_2007}. In that case, dynamics may be accounted for through an action that can be identified as analogous to a Gibbs measure for a system at equilibrium. This latter can be sampled by using a Markov chain Monte-Carlo  algorithm \cite{duben_monte_2008,mesterhazy_anomalous_2011,mesterhazy_lattice_2013} that provides a framework for rare-event sampling.
%
An alternative approach is based on \emph{instanton} \cite{grafke_instanton_2015}. For stochastically driven systems in the limit of weak noise, it can be proved that the probability of extreme events is dominated by the minima of the action \cite{freidlin2013random} -- the trajectories that achieve such minima being referred to as instantons \cite{gurarie_instantons_1996}. Therefore, suitable numerical schemes can  be used to minimize the action and evaluate instantons as well as the related probabilities of rare events \cite{chernykh_large_2001,grafke_instanton_2013,grafke_relevance_2015,grigorio_instantons_2017}.
%
An obvious drawback of the aforementioned approaches is their limitation to stochastically driven dynamics which, in addition, must be simple enough so that an action may be derived and sampled, or minimized. 
%
%
%Whilst Burgers' turbulence provides an interesting model framework for the investigation of the properties of small scale turbulence, it is of little relevance to more realistic situations, such as industrial or environmental flows.
%
In this paper, a more general approach is investigated. It relies on sampling algorithms connected to the simulation of the actual dynamics itself, \textit{i.e.} applying directly to the trajectories in the dynamic space.
% dedicated to dynamical systems for which an action cannot be easily derived or minimized,
%    
%
Even though such ideas trace back to the early 1950's, they have received ever-growing interest over the last 20 years with successful applications in various domains such as chemistry \cite{van_erp_elaborating_2005,escobedo_transition_2009,teo_adaptive_2016}, biophysics \cite{huber_weighted-ensemble_1996,zuckerman2017weighted,bolhuis2005kinetic}, nuclear physics \cite{louvin2017}, nonlinear dynamical systems \cite{tailleur_probing_2007} or communication networks simulation \cite{villen-altamirano_restart:_1994}.
%
% already mentioned in the beginning? This sentence is misleading.
%
%However, a large body of algorithms are limited to systems at equilibrium, for which the stationary measure can be sampled by means of Monte Carlo methods. 
%
An original contribution of the present study is certainly the application of rare-event sampling algorithms in the context of far-from-equilibrium dynamics with an irreducible very large number of degrees of freedom. 
Let us remark that most of previous successful applications involve dynamical systems at statistical equilibrium and low-dimensional attractors. \EL{is it TRUE? \#11}
% 
Two different algorithms suitable for out-of-equilibrium dynamics are considered and compared. Namely, the Adaptive Multilevel Splitting algorithm and the Giardina-Kurchan-Tailleur-Lecomte algorithm.
%

@@

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{ecoulement_typique/ecoulement_typique.png}
	\caption{Snapshots of the vorticity related to typical drag fluctuations (within one standard deviation) over a time interval of length $\tau_c \simeq 4\tau_0$; $\tau_c$ will be identified later as the correlation time of the drag signal.
	The vorticity is given in lattice units.}
	\label{fig:typical_vorticity}
\end{figure}

@@

\begin{figure}
	\centering
	%\includegraphics[width=\linewidth]{timeseries_extrms_AVG/timeseries_extrms_AVG}
	\caption{Instantaneous drag signals $f_d(t)$ corresponding to the highest fluctuations of the time-averaged drag $F_T$ with averaging period $T=10~\tau_c$ and $\tau_c$ being the correlation time of the drag. $\sigma_T$ denotes the standard deviation of $F_T$. \EL{FIGURE IS MISSING} \TL{Addressed in issue \#9}}
	\label{fig:extreme_avg}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\linewidth]{illustr_AMS/figure_AMS}
  \caption{\label{fig:illustr_AMS} Selection-mutation procedure of the \ac{ams} algorithm to compute the probability that a real-valued observable $A(\mathbf{x}(t))$ reaches values larger than $Q$ throughout a trajectory of duration $T_a$.\EL{QUID of the score function mentioned in the text: is the observable? \#1} The initial ensemble of blue trajectories (associated with the time evolution of the observable $A$) is first generated. These trajectories are independent and can be simulated in parallel. On the basis of the respective maxima $Q_1$, $Q_2$ and $Q_3$, the trajectory with the lowest maximum is discarded in the ensemble (dashed blue line). Among the two remaining trajectories, trajectory 3 is chosen randomly and copied until it reaches the value $Q_1$. It is then freely simulated from the branching point to the final time $T_a$. In case of deterministic dynamics, a small perturbation is introduced at the branching to separate the trajectories. This procedure is iterated $J$ times or until all trajectories exceed a fixed threshold $Q$.}
\end{figure}

@@

% Application to instantaneous drag
%
% Can the \ac{tams} algorithm achieve similar performance when applied to complex chaotic dynamics, such as turbulent flows around objects ?
% To answer this question, we consider again the two-dimensional test flow introduced in section~\ref{sec:test_flow}.
%
Our aim is now to sample flow evolution that exhibit extreme fluctuations of the drag $f_d$ acting on the square obstacle.
In contrast with the \acl{ou} process, the phase space is here higlhy-dimensional with intrinsically chaotic dynamics.
In that situation, the choice of the score function $\xi (\mathbf{x}(t),t)$ is no longer straightforward.
However, a natural simple choice for which we will opt in this study is certainly $\xi (\mathbf{x}(t),t) = f_d(t)$. 
\TL{Let us stress here that the flow dynamics are deterministic.
As a result, resampled trajectories do not differ from the parent trajectory their are resampled
from.
A workaround is to introduce a slight perturbation in the state of the flow at the branching point (see figure~\ref{fig:illustr_AMS}).
Because of the flow chaoticity, resampled trajectories will separate from their parent over the \textit{Lyapunov timescale} $\tau_L$.}\EL{Freddy: provide reference for Lyapunov timescale. \#12}

@@

The difference between the typical duration of drag fluctuations $\tau_c$ and the Lyapunov timescale $\tau_L$ may be heuristically associated with the ``convective nature'' of the flow.
%
As discussed previously, the duration of extreme fluctuations of the drag are closely related to the sweeping time of the flow past the obstacle, and therefore to the mean velocity $U$. On the contrary, the Lyapunov timescale is rather associated with the intrinsic evolution of turbulent fluctuations in the reference frame of the mean flow, \textit{i.e.} with the fluctuating velocity $u_{rms}$. The ratio $u_{rms}/U < 1$ in our case of grid-generated turbulence implies that $\tau_L > \tau_c$. \EL{does it make sense to you?} \TL{A priori oui. Mais j'aimerai prendre 10-15min pour bien y réféchir \#2}

@@

The sampling of extreme fluctuations of the time-averaged drag $F_T$ is now examined.
The \ac{tams} algorithm could be used in the same way as before by choosing $F_T$ as score function. However, this would lead to the same pitfall.
% the re-sampling fails to efficiently generate new higher fluctuations of $F_T$.
%
For a time-averaged observable,  an alternative approach is provided by the \acf{gktl} algorithm \cite{giardina_direct_2006,tailleur_probing_2007,giardina_simulating_2011}.
%
Similarly to the \ac{tams} algorithm, the \ac{gktl} algorithm relies on the simulation of an ensemble of trajectories.
In this case, however, the trajectories interact dynamically. 
%
At regular time intervals, some elements of the ensemble are killed and others are cloned according to a weight that depends on the history of the replica.
%
The weights are chosen so that, after several iterations of the algorithm, the trajectories generated are distributed according to a biased probability distribution in favor of trajectories with large values of the (time-averaged) observable of interest.
%
The \ac{gktl} algorithm belongs to a family of algorithms called ``go with the winners'' \cite{aldous1994go,grassberger2002go}.
Similar ideas have been applied in a wide range of fields over the past 50 years, under different names \EL{donner des exemples \#3}, depending of the specific application domain~\cite{moral_feynman-kac_2004}.
The application of a go-with-the-winners approach to the computation of large deviations in non-equilibrium systems has first been proposed in 2006 \cite{giardina_direct_2006}.
Over the past ten years, it has successfully been applied to investigate rare events in both stochastic \cite{giardina_direct_2006,lecomte_numerical_2007,garrahan2007dynamical} and deterministic systems \cite{giardina_direct_2006,tailleur_probing_2007}.
% It was originally designed to compute large deviations rate functions of time-averaged dynamical observables, that describe the probability density of both typical of rare events in the limit of large
% averaging time~\cite{ref_large_dev}.
% By contrast with the \ac{ams} and \ac{tams}, the \ac{gktl} algorithm does not follow the strategy of splitting algorithms.
% Instead, it implements \textit{importance sampling}, \textit{i.e.} sampling a modified distribution that is biased towards the rare events of interest.
% The idea of importance sampling is very general and was used in many different contexts (see e.g.~\cite{Berg1992,Hartmann2002} and the general references~\cite{Bucklew2004,RubinoTuffin2009}). 
% The \ac{gktl} algorithm performs importance sampling in the space of trajectories, which is relevant for out-of-equilibrium systems.

@@

Let 
$\mathbb{P}_{0}\left(\left\{ \mathbf{X}(t)\right\} _{0\leq t\leq T_{a}} = \left\{ \mathbf{x}(t)\right\} _{0\leq t\leq T_{a}}\right)$ 
refer formally to the probability of the trajectory 
$\left\{ \mathbf{x}(t)\right\} _{0\leq t\leq T_{a}}$ of our dynamical system.
%
% $\mathbb{P}_{k}\left(\left\{ \mathbf{X}(t)\right\} _{0\leq t\leq T_{a}} = \left\{ \mathbf{x}(t)\right\} _{0\leq t\leq T_{a}} \right)$ 
%
By construction, the probability to obtain the same trajectory by resorting to the previous cloning procedure is given by
%
\begin{align}
\mathbb{P}_{k}\left(\left\{ \mathbf{X}(t)\right\} _{0\leq t\leq T_{a}}=\left\{ \mathbf{x}(t)\right\} _{0\leq t\leq T_{a}}\right) &\underset{N\rightarrow\infty}{\sim} \frac{e^{k\int_{0}^{T_{a}}A(\mathbf{x}(t))dt}}{Z(k,T_a)}\mathbb{\mathbb{P}}_{0}\left(\left\{ \mathbf{X}(t)\right\} _{0\leq t\leq T_{a}}=\left\{ \mathbf{x}(t)\right\} _{0\leq t\leq T_{a}}\right)
\label{eq:Biased_Path_Approximation}
\end{align}
where the normalization factor is given by $Z(k,T_a)=\mathbb{E}_{0}\left[e^{k\int_{0}^{T_{a}}A(\mathbf{X}(t))dt}\right]$ with $\mathbb{E}_{0}$ being the expectation value with respect to $\mathbb{P}_{0}$.
%
% obvious? 
%and $\underset{N\rightarrow\infty}{\sim}$ meaning that this is true only asymptotically for large $N$.
%
%
%The typical error is of order $1/\sqrt{N}$ when evaluating averages over observables.
%
Eq.~\eqref{eq:Biased_Path_Approximation} relies on the mean-field approximation
\begin{equation}
R_{1}=\frac{1}{N}\sum_{n=1}^{N}e^{k\int_{t_0}^{t_{1}}A(\mathbf{X}_{n}(t))dt}\underset{N\rightarrow\infty}{\sim} Z(k,t_1)= \mathbb{E}_{0}\left[e^{k\int_{t_0}^{t_{1}}A(\mathbf{X}(t))dt}\right]
\label{eq:Mean_Field_Approximation}
\end{equation}
The typical relative error related to this approximation can be shown of being of order $1/\sqrt{N}$\cite{DelMoralBook,DelMoral2013} for a family of rare-event algorithms including the \ac{gktl} algorithm.
which by induction (and assuming similar relations for the successive $R_i$'s) eventually leads  to~\cite{giardina_direct_2006,giardina_simulating_2011}
\begin{equation}
\prod_{i=1}^{T_{a}/\tau}R_{i}\underset{N\rightarrow\infty}{\sim} Z(k,T_a) =\mathbb{E}_{0}\left[e^{k\int_{0}^{T_a}A(\mathbf{X}(t))dt}\right].\label{eq:Estimate_Lambda}
\end{equation}
\EL{IS IT TRUE?} \TL{Oui. Il faudrait un peu reécrire ce paragraphe pour que ca soit clair, voir issue \#5}
The validity of this mean-field approximation and the 
%
%
The killed trajectories, \emph{i.e.} with a weight smaller than $1$ \EL{true?}, are discarded from the statistics.
%
% no because the trajectories are cloned
%
% Starting from the final $N$ trajectories at time $T_{a}$, one goes backwards in time through the selection events attaching to each piece of trajectory its ancestor.
%
Eventually, an effective ensemble of $N$ trajectories for $0 \leq t \leq T_{a}$ is obtained, distributed according to $\mathbb{P}_{k}$.
All trajectories are solutions of the dynamical system since we have not modified the dynamics, but only sampled trajectories according to the distribution $\mathbb{P}_{k}$ rather than according to the distribution $\mathbb{P}_{0}$.
\EL{quid the small perturbation to separated the clones? comment required \#6}
%
The sampled trajectories can be used to compute the statistical properties of any observable with respect to the distribution $\mathbb{P}_{0}$ from the distribution $\mathbb{P}_{k}$ by using Eq.~\eqref{eq:Biased_Path_Approximation}.
%
More details concerning the implementation of the algorithm and the corresponding data analysis are provided in \cite{lestang:tel-01974316,brewer2018efficient}.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{IS_GKTL/IS_GKTL}
	\caption{\label{fig:IS_GKTL} Rare-event sampling of the (zero-mean) time-averaged drag $\tilde F_T = F_T - \overline{F_T}$ with $T=10~\tau_c$; $\tau_c$ is the correlation time of the instantaneous drag. The shaded \ac{pdf}s are estimated from the biased ensemble resulting from the \ac{gktl} algorithm applied to $16384$ trajectories of duration $T_a=T$ with the cloning period $\tau=\tau_c/2$.
	The dashed line denotes the unbiased \ac{pdf} of $\tilde F_T$, \textit{i.e.} from direct sampling or with $k=0$ (no bias) in the \ac{gktl} algorithm.
\EL{$F_T-\overline{F_T}$ in the figure! \#4}}
\end{figure}

@@

\EL{rajouter le calcul des temps de retour par gktl pour etoffer un peu la partie du papier sur gktl -- il faudrait montrer un resultat qui montre que l'on obtient la bonne statistique} \TL{Demande pas mal de travail en comparaison des autres points. Voir issue \#10.}

@@

\EL{CONCLUSION TO BE REVIEWED \#7}
